---
title: "| OPERATIONS MANUAL\n| \n| Strategus Data Model\n| \n| \n"
author: "Marc A. Suchard"
fontsize: 12pt
geometry: margin=1in
always_allow_html: true
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    highlight: zenburn
    css: "style.css"
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    # citation_package: natbib
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
params:
  date: '2023-06-01'
  version: 0.0.99
subtitle: 'Version: `r params$version`'
link-citations: true
csl: bmj.csl
---

```{r setup, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(kableExtra)
library(DT)

latex_table_font_size <- 8
```

# Access and schema

This document describes the data models for `Strategus` modules and some instructions on how to access results in these models.
We hold results on the OHDSI public database server `shinydb` that runs `postgresql`.
You can install the `postgresql` JDBC drivers; this only needs to be done once.

It is recommend to specify the environmental variable `DATABASECONNECT_JAR_FOLDER=<folder-of-your-choice>` in `.Renviron` located in the user's home directory.
In the example below, however, we set the environmental variable manually in `R` to make this document more self-contained.

```{r db, eval=FALSE}
Sys.setenv(DATABASECONNECTOR_JAR_FOLDER = getwd())
DatabaseConnector::downloadJdbcDrivers(dbms = "postgresql")
```

and set-up connection details, in this case using `keyring` to hide usernames and passwords for security`

```{r connect, eval=TRUE}
# OHDSI shinydb read-only credentials
connectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = keyring::key_get("sosTutorialServer"),
  user = keyring::key_get("sosTutorialUser"),
  password = keyring::key_get("sosTutorialPassword"))
```

Table \@ref(tab:schema) provides the `schema` names for SOS Challenge results.

```{r schema, echo=FALSE, warning=FALSE}
schema <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, 
                            show_col_types = FALSE,file = "
  Fluoroquinolone and aortic aneurysms ; quinoloneaa
  Intravitreal anti-VEGF and kidney failure ; antivegfkf
")

tab <- kbl(schema, col.names = c("SOS Challenge study", "schema"),
           linesep = "", booktabs = TRUE,
           caption = "SOS Challenge study schema")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped",
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Data models

```{r echo=FALSE}
printDataModelSpecification <- function(filename, level = "###") {

  specifications <- readr::read_csv(filename, show_col_types = FALSE) 
  
  if ("optional" %in% names(specifications)) {
    specifications <- specifications %>% filter(optional == "No")
  }
  
  tables <- split(specifications, specifications$table_name)
  
  for (table in tables) {
    header <- sprintf("%s Table %s", level, table$table_name[1])
    
    table <- table %>%
      select(Field = .data$column_name, Type = .data$data_type, Key = .data$primary_key
             # , Description = .data$description
      ) %>%
      kbl(linesep = "", booktabs = TRUE, longtable = TRUE)
    
    if (knitr::is_latex_output()) {
      writeLines("")
      writeLines(header)
      
      writeLines(table %>%
                   kable_styling(latex_options = "striped", font_size = latex_table_font_size) %>%
                   column_spec(1, width = "10em") %>%
                   column_spec(2, width = "5em") %>%
                   column_spec(3, width = "3em") %>%
                   column_spec(4, width = "16em"))
    } else if (knitr::is_html_output()) {
      writeLines("")
      writeLines(header)
      
      writeLines(table %>%
                   kable_styling(bootstrap_options = "striped"))
    }
  }
}
```

```{r echo=FALSE, results="asis", warning=FALSE, cache=TRUE}
modules <- readr::read_delim(col_names = TRUE, delim = ";", trim_ws = TRUE, 
                            show_col_types = FALSE,file = "
  name ; prefix ; url
  Characterization ; c ; https://raw.githubusercontent.com/OHDSI/Characterization/main/inst/settings/resultsDataModelSpecification.csv
  CohortDiagnostics ; cd ; https://raw.githubusercontent.com/OHDSI/CohortDiagnostics/main/inst/settings/resultsDataModelSpecification.csv
  CohortGenerator; gc ; https://raw.githubusercontent.com/OHDSI/CohortGeneratorModule/main/resultsDataModelSpecification.csv
  CohortIncidence ; ci; https://raw.githubusercontent.com/OHDSI/CohortIncidenceModule/master/resultsDataModelSpecification.csv
  CohortMethod ; ; https://raw.githubusercontent.com/OHDSI/CohortMethod/main/inst/csv/resultsDataModelSpecification.csv
  PatientLevelPrediction ; plp ; https://raw.githubusercontent.com/OHDSI/PatientLevelPrediction/main/inst/settings/resultsDataModelSpecification.csv
")

for (i in 1:nrow(modules)) {
  cat("## ", modules[i,]$name)
  printDataModelSpecification(filename = modules[i,]$url,
                              level = "###")
}
```

## Fields with minimum values

Some fields contain patient counts or fractions that are easily converted to patient counts.
To prevent identifiability, these fields are subject to a minimum value.
When the value falls below this minimum, it is replaced with the negative value of the minimum.
For example, if the minimum subject count is 5, and the actual count is 2, the value stored in the data model will be -5, which could be represented as '\<5' to the user.
Note that the value 0 is permissible, as it identifies no persons.

# Retrieving database meta-data

We will use the `Hades` packages `DatabaseConnector` and `SqlRender` to query the results tables directly in `R`.  
Many `join`s and column-`select` can be done either on the server-side (`postgresql`) or client-side (`R`).
We prefer the server-side when accessing large tables.
On the client-side, `dplyr` is effective.

Here is an example of a basic query to `database_meta_data` to return data-source names and abbreviations (with client-side `select`):

```{r, metaData, message=FALSE, cache=TRUE}
schema <- "quinoloneaa"

connection <- DatabaseConnector::connect(connectionDetails = connectionDetails)

sql <- SqlRender::render(sql = "
  SELECT * FROM @schema.database_meta_data",
  schema = schema) # parameterized
                        
metaData <- DatabaseConnector::querySql(connection = connection,
                                        sql = sql,
                                        snakeCaseToCamelCase = TRUE)

DatabaseConnector::disconnect(connection)

names(metaData)

library(dplyr)
databaseName <- metaData %>% select(cdmSourceAbbreviation, databaseId)
databaseName
```

Now is a good time to (manually) change the abbreviations in `databaseName` to consistent, similar-length, reader-friendly choices for our manuscript(s).

```{r}
databaseName$shortDatabaseName <- c("JMDC", "MDCD", "Ausom", "NHIS", "OptumDOD", 
                                    "VA", "OptumEHR", "CCAE", "PhM", "TMU")
```

Likewise, we retrieve reader-friendly cohort names.
Note that the following codes does not return negative control cohort names; these are not (?) currently stored in the results-set.

```{r, cohortNames, message=FALSE, cache=TRUE}
connection <- DatabaseConnector::connect(connectionDetails = connectionDetails)
sql <- SqlRender::render(sql = "
  SELECT DISTINCT cohort_id, cohort_name FROM @schema.cg_cohort_generation",
  schema = schema)
                        
cohortName <- DatabaseConnector::querySql(connection = connection,
                                            sql = sql,
                                            snakeCaseToCamelCase = TRUE) 

DatabaseConnector::disconnect(connection)
```

Often these names are unusable for reader-friendly documents.
See, e.g., `cohortName[6, "cohortName"]` which is `r cohortName[1, "cohortName"]`.
We will re-code these names for prettier output.
Here we demonstrate `DT::datatable` that provides an interactive (HTML-only) table renderer useful for web-pages and presentations.

```{r}
cohortName$shortCohortName <- c("AD", "UTI", "Cef", "AA", "FQ", "FQ2", 
                                "UTI2", "Cef2", "TMP", "AA/AD", "TMP2")

DT::datatable(cohortName)
```

# Basic results queries

Here is an example of a basic query to a `cohort_counts` table to return cohort sizes.
We then output the counts using a well-formatted table

```{r, cohortCounts, message=FALSE, cache=TRUE}
connection <- DatabaseConnector::connect(connectionDetails = connectionDetails)
sql <- SqlRender::render(sql = "
  SELECT database_id, cohort_id, cohort_subjects FROM @schema.cg_cohort_count",
  schema = schema)
                        
cohortCount <- DatabaseConnector::querySql(connection = connection,
                                            sql = sql,
                                            snakeCaseToCamelCase = TRUE) %>% 
  inner_join(databaseName, by = "databaseId") %>% 
  inner_join(cohortName, by = "cohortId") %>%
  select(shortCohortName, shortDatabaseName, cohortSubjects) %>%
  rename(cohortName = shortCohortName, databaseName = shortDatabaseName) %>%
  arrange(cohortName,  databaseName)

DatabaseConnector::disconnect(connection)
```

## Hints for beautiful tables

The `kableExtra` package (especially when printed to PDF) makes beautiful tables.

```{r, message=FALSE}
table <- cohortCount %>% filter(cohortName %in% c("FQ", "TMP")) %>%
  rename(Cohort = cohortName, Database = databaseName, Subjects = cohortSubjects) %>%
  tidyr::spread(Cohort, Subjects)

tab <- table %>% kbl() %>% kable_classic(full_width = F) %>%
  add_header_above(c(" " = 1, "Exposure" = 2)) %>%
  row_spec(0, align = "c")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped",
                          font_size = latex_table_font_size)
} else {
  tab ## bootstrap-striped is not nicely compatible with `add_header_above`
}
```

Often it is most convenient to hand-craft `latex` table headers and footers and `cat` this material directly to the output.
Here is a silly example.


```{r, echo=TRUE, eval=FALSE}
# Requires "results='asis'"

library(xtable)
cat(gsub("\\\\", "\\", r"{
  \begin{table}[h]
  \begin{center}
  \begin{tabular}{lrr}
  \hline
  \multicolumn{1}{c}{$\alpha$} & \multicolumn{1}{c}{$\beta$} & \multicolumn{1}{c}{$\delta$} \\\\
}", fixed = TRUE))

print(xtable(table, format = "latex"), 
      include.rownames = FALSE, include.colnames = FALSE, only.contents = TRUE)

cat(gsub("\\\\", "\\", r"{
  \end{tabular}
  \end{center}
  \caption{A silly caption with \LaTeX commands}
  \end{table}
}", fixed = TRUE))
```

```{r, echo=FALSE, results='asis', message=FALSE}
# Requires "results='asis'"

library(xtable)
cat(gsub("\\\\", "\\", r"{
  \begin{table}[h]
  \begin{center}
  \begin{tabular}{lrr}
  \hline
  \multicolumn{1}{c}{$\alpha$} & \multicolumn{1}{c}{$\beta$} & \multicolumn{1}{c}{$\delta$} \\\\
}", fixed = TRUE))

print(xtable(table, format = "latex"), 
      include.rownames = FALSE, include.colnames = FALSE, only.contents = TRUE)

cat(gsub("\\\\", "\\", r"{
  \end{tabular}
  \end{center}
  \caption{A silly caption with \LaTeX commands}
  \end{table}
}", fixed = TRUE))

```

Note that if one abhors escaping backslash-characters, then one can read the header and footer from text files.

# Working with `CohortMethod` results

Unfortunately, we currently do *not* have a shared repository of functions to generate typical results figures and tables.
Even `OhdsiShinyModules` hides these tools inside inaccessible `shiny`-functions.
Until we solve this limitation, we recommend using copying code from previous public manuscripts.
The `LegendT2dmEvidenceExplorer` from LEGEND-T2DM is a good example: (`getCovariateBalance()`)[https://github.com/ohdsi-studies/LegendT2dmEvidenceExplorer/blob/b3564ea3c8257817d92de24b0f2244fed1df1f99/inst/shiny/DataPulls.R#L331]
 


[ADD EXAMPLE]

