---
title: "| SOS Challenge \n| \n| Evidence Tutorial #1\n| \n| \n"
author: "Marc A. Suchard"
fontsize: 12pt
geometry: margin=1in
always_allow_html: true
output:
  bookdown::html_document2:
    df_print: paged
    toc: yes
    toc_depth: 2
    toc_float: yes
    number_sections: yes
    number_tables: yes
    highlight: zenburn
    css: "style.css"
  bookdown::pdf_document2:
    keep_tex: yes
    latex_engine: xelatex
    md_extensions: +raw_attribute
    number_sections: yes
    # citation_package: natbib
    includes:
      before_body: title.tex
header-includes:
- \usepackage[numbers,sort&compress]{natbib}
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage[normalem]{ulem}
- \usepackage{makecell}
- \usepackage{caption}
- \usepackage{rotating}
- \usepackage{multirow}
- \usepackage{mwe,tikz}
- \usepackage[percent]{overpic}
- \usepackage{enumitem}
- \usepackage{hyperref}
- \newcolumntype{P}[1]{>{\raggedright\arraybackslash}p{#1}}
- \newcommand{\footerDate}{`r params$date`}
- \input{header.tex}
longtable: yes
mainfont: Arial
params:
  date: '2023-06-01'
  version: 0.0.99
subtitle: 'Version: `r params$version`'
link-citations: true
csl: bmj.csl
---

```{r setup, include=FALSE}
def.chunk.hook  <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(chunk = function(x, options) {
  x <- def.chunk.hook(x, options)
  paste0("\n \\", "footnotesize","\n\n", x, "\n\n \\normalsize")
})
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(kableExtra)
library(DT)

latex_table_font_size <- 8

Sys.setenv(TESTTHAT = "TRUE")
```

# Installation

Follow the instructions on the HADES website to [how to set up R](https://ohdsi.github.io/Hades/rSetup.html) and [how to install HADES](https://ohdsi.github.io/Hades/installingHades.html).

While this tutorial does not use many of the HADES packages, these are often quite useful to have.  
This tutorial does require `DatabaseConnector` and `SqlRender` from HADES, along with `keyring`, dplyr`, `kableExtra`, `DT` and their dependencies.
These can be installed directly from CRAN.

```{r install, eval=FALSE}
install.packages(c("SqlRender", "DatabaseConnector",
                   "keyring", "dplyr", "kableExtra"))
```

We will use `keyring` to hide database credentials.
The following codes only needs to be executed once per machine.

```{r credentials, eval=FALSE}
keyring::key_set_with_value("sosTutorialUser", password = "<FILL-IN>")
keyring::key_set_with_value("sosTutorialPassword", password = "<FILL-IN>")
keyring::key_set_with_value("sosTutorialServer", password = "<FILL-IN>")
```

# Access and schema

This document describes the data models for `Strategus` modules and some instructions on how to access results in these models.
We hold results on the OHDSI public database server `shinydb` that runs `postgresql`.
You can install the `postgresql` JDBC drivers; this only needs to be done once.

It is recommend to specify the environmental variable `DATABASECONNECT_JAR_FOLDER=<folder-of-your-choice>` in `.Renviron` located in the user's home directory.
In the example below, however, we set the environmental variable manually in `R` to make this document more self-contained.

```{r db, eval=FALSE}
Sys.setenv(DATABASECONNECTOR_JAR_FOLDER = getwd())
DatabaseConnector::downloadJdbcDrivers(dbms = "postgresql")
```

and set-up connection details, in this case using `keyring` to hide usernames and passwords for security`

```{r connect, eval=TRUE}
# OHDSI shinydb read-only credentials
connectionDetails <- DatabaseConnector::createConnectionDetails(
  dbms = "postgresql",
  server = keyring::key_get("sosTutorialServer"),
  user = keyring::key_get("sosTutorialUser"),
  password = keyring::key_get("sosTutorialPassword"))
```

Table \@ref(tab:schema) provides the `schema` names for SOS Challenge results.

```{r schema, echo=FALSE, warning=FALSE}
schema <- readr::read_delim(col_names = FALSE, delim = ";", trim_ws = TRUE, 
                            show_col_types = FALSE,file = "
  Fluoroquinolone and aortic aneurysms ; quinoloneaa
  Intravitreal anti-VEGF and kidney failure ; antivegfkf
")

tab <- kbl(schema, col.names = c("SOS Challenge study", "schema"),
           linesep = "", booktabs = TRUE,
           caption = "SOS Challenge study schema")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped",
                        font_size = latex_table_font_size)
} else {
  tab %>% kable_styling(bootstrap_options = "striped")
}
```

# Retrieving database meta-data

We will use the `Hades` packages `DatabaseConnector` and `SqlRender` to query the results tables directly in `R`.  
Many `join`s and column-`select` can be done either on the server-side (`postgresql`) or client-side (`R`).
We prefer the server-side when accessing large tables.
On the client-side, `dplyr` is effective.

Here is an example of a basic query to `database_meta_data` to return data-source names and abbreviations (with client-side `select`):

```{r, metaData, message=FALSE, cache=TRUE}
schema <- "quinoloneaa"

connection <- DatabaseConnector::connect(connectionDetails = connectionDetails)

sql <- SqlRender::render(sql = "
  SELECT * FROM @schema.database_meta_data",
  schema = schema) # parameterized
                        
metaData <- DatabaseConnector::querySql(connection = connection,
                                        sql = sql,
                                        snakeCaseToCamelCase = TRUE)

DatabaseConnector::disconnect(connection)

names(metaData)

library(dplyr)
databaseName <- metaData %>% select(cdmSourceAbbreviation, databaseId)
databaseName
```

Now is a good time to (manually) change the abbreviations in `databaseName` to consistent, similar-length, reader-friendly choices for our manuscript(s).

```{r}
databaseName$shortDatabaseName <- c("JMDC", "MDCD", "Ausom", "NHIS", "OptumDOD", 
                                    "VA", "OptumEHR", "CCAE", "PhM", "TMU")
```

Likewise, we retrieve reader-friendly cohort names.
Note that the following codes does not return negative control cohort names; these are not (?) currently stored in the results-set.

```{r, cohortNames, message=FALSE, cache=TRUE}
connection <- DatabaseConnector::connect(connectionDetails = connectionDetails)
sql <- SqlRender::render(sql = "
  SELECT DISTINCT cohort_id, cohort_name FROM @schema.cg_cohort_generation",
  schema = schema)
                        
cohortName <- DatabaseConnector::querySql(connection = connection,
                                            sql = sql,
                                            snakeCaseToCamelCase = TRUE) 

DatabaseConnector::disconnect(connection)
```

Often these names are unusable for reader-friendly documents.
See, e.g., `cohortName[6, "cohortName"]` which is `r cohortName[1, "cohortName"]`.
We will re-code these names for prettier output.
Here we demonstrate `DT::datatable` that provides an interactive (HTML-only) table renderer useful for web-pages and presentations.

```{r}
cohortName$shortCohortName <- c("AD", "UTI", "Cef", "AA", "FQ", "FQ2", 
                                "UTI2", "Cef2", "TMP", "AA/AD", "TMP2")

DT::datatable(cohortName)
```

# Basic results queries

Here is an example of a basic query to a `cohort_counts` table to return cohort sizes.
We then output the counts using a well-formatted table

```{r, cohortCounts, message=FALSE, cache=TRUE}
connection <- DatabaseConnector::connect(connectionDetails = connectionDetails)
sql <- SqlRender::render(sql = "
  SELECT database_id, cohort_id, cohort_subjects FROM @schema.cg_cohort_count",
  schema = schema)
                        
cohortCount <- DatabaseConnector::querySql(connection = connection,
                                            sql = sql,
                                            snakeCaseToCamelCase = TRUE) %>% 
  inner_join(databaseName, by = "databaseId") %>% 
  inner_join(cohortName, by = "cohortId") %>%
  select(shortCohortName, shortDatabaseName, cohortSubjects) %>%
  rename(cohortName = shortCohortName, databaseName = shortDatabaseName) %>%
  arrange(cohortName,  databaseName)

DatabaseConnector::disconnect(connection)
```

## Hints for beautiful tables

The `kableExtra` package (especially when printed to PDF) makes beautiful tables.

```{r, message=FALSE}
table <- cohortCount %>% filter(cohortName %in% c("FQ", "TMP")) %>%
  rename(Cohort = cohortName, Database = databaseName, Subjects = cohortSubjects) %>%
  mutate(Subjects = prettyNum(Subjects, 
                              format = "d", 
                              big.mark = ",")) %>% # Can easily change for _Lancet_
  tidyr::spread(Cohort, Subjects)

tab <- table %>% kbl(align = "lrr") %>% kable_classic(full_width = F) %>%
  add_header_above(c(" " = 1, "Exposure" = 2)) %>%
  row_spec(0, align = "c")

if (knitr::is_latex_output()) {
  tab %>% kable_styling(latex_options = "striped",
                          font_size = latex_table_font_size)
} else {
  tab ## bootstrap-striped is not nicely compatible with `add_header_above`
}
```

Often it is most convenient to hand-craft `latex` table headers and footers and `cat` this material directly to the output.
Here is a silly example.


```{r, echo=TRUE, eval=FALSE}
# Requires "results='asis'"

library(xtable)
cat(gsub("\\\\", "\\", r"{
  \begin{table}[h]
  \begin{center}
  \begin{tabular}{lrr}
  \hline
  \multicolumn{1}{c}{$\alpha$} & \multicolumn{1}{c}{$\beta$} & \multicolumn{1}{c}{$\delta$} \\\\
}", fixed = TRUE))

print(xtable(table, format = "latex"), 
      include.rownames = FALSE, include.colnames = FALSE, only.contents = TRUE)

cat(gsub("\\\\", "\\", r"{
  \end{tabular}
  \end{center}
  \caption{A silly caption with \LaTeX commands}
  \end{table}
}", fixed = TRUE))
```

```{r, echo=FALSE, results='asis', message=FALSE}
library(xtable)
cat(gsub("\\\\", "\\", r"{
  \begin{table}[h]
  \begin{center}
  \begin{tabular}{lrr}
  \hline
  \multicolumn{1}{c}{$\alpha$} & \multicolumn{1}{c}{$\beta$} & \multicolumn{1}{c}{$\delta$} \\\\
}", fixed = TRUE))

print(xtable(table, format = "latex"), 
      include.rownames = FALSE, include.colnames = FALSE, only.contents = TRUE)

cat(gsub("\\\\", "\\", r"{
  \end{tabular}
  \end{center}
  \caption{A silly caption with \LaTeX commands}
  \end{table}
}", fixed = TRUE))

```

Note that if one abhors escaping backslash-characters, then one can read the header and footer from text files.

# Working with `CohortMethod`

Unfortunately, we currently do *not* have a shared repository of functions to generate typical results figures and tables.
Even `OhdsiShinyModules` hides these tools inside inaccessible `shiny`-functions.
Until we solve this limitation, we recommend duplicating code from previous public manuscripts.
The `LegendT2dmEvidenceExplorer` from LEGEND-T2DM is a good example: [`getCovariateBalance()`](https://github.com/ohdsi-studies/LegendT2dmEvidenceExplorer/blob/b3564ea3c8257817d92de24b0f2244fed1df1f99/inst/shiny/DataPulls.R#L331)
 
## Query the design choices and comparisons

```{r, cache=TRUE, message=FALSE, warning=FALSE}
connection <- DatabaseConnector::connect(connectionDetails)

sql <- paste0("SET search_path TO ", schema, ";")
DatabaseConnector::executeSql(connection = connection, sql = sql)

sql <- "
  SELECT DISTINCT target_id, comparator_id,
    cm_analysis.analysis_id, cm_analysis.description
  FROM cm_covariate_balance
  INNER JOIN cm_analysis
  ON cm_analysis.analysis_id = cm_covariate_balance.analysis_id"

analysis <- DatabaseConnector::querySql(connection = connection,
                                        sql = sql,
                                        snakeCaseToCamelCase = TRUE)

DatabaseConnector::disconnect(connection)

datatable(analysis %>% arrange(analysisId, targetId, comparatorId))
```

## Covariate balance

```{r, cache=TRUE}
# Stolen from `ShinyShinyModules`
getCovariateBalance <- function(connection, schema,
                                targetId,
                                comparatorId,
                                analysisId,
                                databaseId,
                                outcomeId) {
  sql <- "  
      SELECT
        cmcb.database_id,
        cmcb.covariate_id,
        cmc.covariate_name,
        cmc.covariate_analysis_id analysis_id,
        cmcb.target_mean_before before_matching_mean_treated,
        cmcb.comparator_mean_before before_matching_mean_comparator,
        abs(cmcb.std_diff_before) abs_before_matching_std_diff,
        cmcb.target_mean_after after_matching_mean_treated,
        cmcb.comparator_mean_after after_matching_mean_comparator,
        abs(cmcb.std_diff_after) abs_after_matching_std_diff
      FROM
        (select * from  @results_schema.@table_prefixcovariate_balance 
        WHERE target_id = @target_id
        AND comparator_id = @comparator_id
        AND outcome_id = @outcome_id
        AND analysis_id = @analysis_id
        AND database_id = '@database_id'
        ) as cmcb
        INNER JOIN @results_schema.@table_prefixcovariate cmc 
          ON cmcb.covariate_id = cmc.covariate_id 
          AND cmcb.analysis_id = cmc.analysis_id 
          AND cmcb.database_id = cmc.database_id"
  
  sql <- SqlRender::render(sql,
                           target_id = targetId,
                           comparator_id = comparatorId,
                           database_id = databaseId,
                           results_schema = schema,
                           table_prefix = "cm_",
                           analysis_id = analysisId,
                           outcome_id = outcomeId)
  
  balance <- DatabaseConnector::querySql(connection, sql,
                                         snakeCaseToCamelCase = TRUE)
  return(balance)
}
```

```{r, cache=TRUE, message=FALSE}
connection <- DatabaseConnector::connect(connectionDetails)

balance <- getCovariateBalance(connection, schema,
                               targetId = 1782488001,     # FQ2
                               comparatorId = 1782487001, # Cef2
                               analysisId = 2,
                               databaseId = -7753527,     # VA
                               outcomeId = 1782489)       # AA/AD

DatabaseConnector::disconnect(connection)

datatable(balance %>% select(-databaseId, -analysisId),
          colnames = c("ID", "Name", "bT", "bC", "bSD", "aT", "aC", "aSD"),
          rownames = NULL)
```

It remains straight-forward to adapt existing plots or generate one's own.

```{r}
plot(balance$absBeforeMatchingStdDiff, balance$absAfterMatchingStdDiff,
     xlim = c(0, 0.4), ylim = c(0, 0.4),
     xlab = "Before", ylab = "After")
abline(h = 0.1, lty = 2)
```

When saving figures to disk, use a vector-graphic format (e.g., PDF), as these formats can scale to any size without resolution loss.
For `R-base` plotting, the relevant functions are `pdf()` and `dev.off()`.
For `gg` plotting, use `ggsave(filename = "<NAME>.pdf")`.

## Effect estimates

